\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-11-10 paper_OJA.tex
more adds
Start mv MVN SVI (sec. 5.3) to Appendix

Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}). As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
function, taking derivatives, and running a gradient-based or coordinate-ascent optimization. First is defined a bijective differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$. One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models. Then, the cost function reads
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-2}
\end{equation}
with
\begin{equation}
p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
\end{equation}
which exhibit the Jacobian of the $T^{-1}$ transformation. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints. \FrL{we should say that before.}

Now, one is faced with the evaluation of the expectations during gradient descent. The solution is to use what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \citep{2015arXiv150603431K}). Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie. a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model. \FrL{Hummm this is a little bit too specific, in practice I would use a flow rather than an MVN.... Maybe this can go to an appendix for people who want to understand how this works on a concrete example?} Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI. As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX. Now even if one can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}. \FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
\FrL{we don't have plots of just the VI?}\JEC{SVI alone (ie. not followed by HMC) gives poor results.}
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

@@

mcmc.run(jax.random.PRNGKey(42))
\end{lstlisting}
%\end{minted}
More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_shear_cluster_ModelReparam_SVI-MVN_Neutra_NUTS}. Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.
%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%\section*{Bibliography}
%\bibliographystyle{elsarticle-harv} 
\bibliographystyle{aa} 
\typeout{}
\bibliography{refs}

>>>>

Using the auto-differentation tool, one can use ``black-box'' guides (aka \textit{automatic differentiation variational inference}). As stated by the authors of \citep{2016arXiv160300788K} ADVI helps specifying a variational family appropriate to the model, computing the corresponding objective
function, taking derivatives, and running a gradient-based or coordinate-ascent optimization. First is defined a invertible differentiable transformation $T$ of the original latent variables $z$ into new variables $\xi$, such $\xi=T(z)$ and $z=T^{-1}(\xi)$. The main advantage of this reparametrisation is that the minimization can be performed with no bound constraints. One can develop ``AutoGuides'' (\numpyro\ terminology) that can be adapted to the user models. Then, the cost function reads
\begin{equation}
\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log q(\xi;\lambda)\right]}_{guide} - \underbrace{\mathbb{E}_{q(\xi;\lambda)}\left[ \log p(\xi,\mathcal{D}) \right]}_{model}
\label{eq-loss-svi-2}
\end{equation}
with
\begin{equation}
p(\xi,\mathcal{D}) \bydef p(T^{-1}(\xi),\mathcal{D}) |J_{T^{-1}}(\xi)|
\end{equation}
which exhibit the Jacobian of the $T^{-1}$ transformation.
\JEC{
The evaluation of the expectations during gradient descent of the loss Eq.~\ref{eq-loss-svi-2} can be done using what is called \textit{elliptical standardisation} or \textit{re-parametrization trick} or \textit{coordinate transformation} (see references in \cite{2015arXiv150603431K}).  Let us illustrated the method using an invertible transformation (aka \textit{normalizing flow})  $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, such that $\zeta\sim \mathcal{N}(0,I)$. The Jacobian of this distribution is 1 by definition (volume conservation), so the lost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the $q$ distribution, so its gradient can be computed once for all for a given $q$ distribution family and reused in any user model.  
Then,
to get $\nabla_\lambda \mathcal{L}$, the $\nabla$ operator can be put inside the expectation which leads to\footnote{To simplify the notation, $T^{-1}(S_\lambda^{-1}(\zeta))$ has been replaced by $z$.}
\begin{multline}
-\nabla_\lambda\mathcal{L}(\lambda) = \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right. \right. \\
+ \left. \left. \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \nabla_\lambda S_\lambda^{-1}(\zeta)
\right\}
+ \nabla_\lambda \mathbb{H}[q(\xi;\lambda)]
\label{eq-loss-svi-4}
\end{multline}
An concrete example of the gradient formula is given in Appendix \ref{app-SVI_MVN} with the multivariate Gaussian distribution family used as the guide.
}%JEC
%

One can sample from the optimized  $q(z,\lambda^\ast)$ distribution, we have used SVI as front-end to a NUTS sampler which is described in the next section \ref{sec-Neural-Reparametrisation} and nicknamed as \textit{Neural Transport}. \FrL{here it would be better to have a small statement on the limitations of VI, and why it's a good idea to then go to Neutra.} 
\FrL{we don't have plots of just the VI?}\JEC{Good point: I have SVI alone with both MVN or Block Neural Autoregressive Flow. The contours are narrower which would require to elaborate...}
%
\subsubsection{Neural Transport}
\label{sec-Neural-Reparametrisation}
%
If the SVI method can be used as is to get $z$ i.i.d. samples from the $q(z,\lambda^\ast)$ distribution, the \textit{Neural Transport MCMC} method \citep{2014arXiv1412.5492P,2019arXiv190303704H} is an efficient way to boost HMC efficiency, especially in target distribution unfavourable geometry where for instance the leapfrog integration algorithm has to face squeezed join distributions for a subset of variables. From SVI, one obtains a first approximation of the target distribution and we show that the change-of-variable $z=F_\lambda(\zeta)$ is such that
\begin{equation}
q(z;\lambda) \rightarrow q(\zeta;\lambda) \bydef q(F_\lambda(\zeta)) |J_{F_\lambda}(\zeta)|
\end{equation}
which is evaluated with optimal $\lambda^\ast$ according to maximization of the ELBO (eg. $F_\lambda=T^{-1}\circ S^{-1}_\lambda$). So, one can use a HMC sampler (eg. NUTS) based on $p(\zeta;\mathcal{D})$ distribution, initialized with $\zeta$ samples from $q(\zeta;\lambda^\ast)$, to get a Markov Chain of $N$ samples $(\zeta_i)_{i<N}$. Then, from the transformation  $z_i=F_{\lambda^\ast}(\zeta_i)$ one finally obtain a Marcov Chain with $(z_i)_{i<N}$ samples. The method is accessible throw the \verb|NeuTraReparam| in \numpyro\ library.

@@

mcmc.run(jax.random.PRNGKey(42))
\end{lstlisting}
%\end{minted}
More details can be found in the repository of this paper to perform a \numpyro\ modeling using \jaxcosmo : a structured NUTS sampling described in Section~\ref{sec-NUTS} \nblink{VectorizedNumPyro_block}, and a SVI optimisation followed by NUTS sampling of Section~\ref{sec-Neural-Reparametrisation} \nblink{DESY_Y1_shear_cluster_ModelReparam_SVI-MVN_Neutra_NUTS}. Notice however that the notebooks are only illustrative and for a real sample production the authors invite the reader to run their own script.

\section{SVI example with MVN guide}
\label{app-SVI_MVN}
\JEC{To be cleaned}
Let us illustrate the trick in the case where $q(\xi; \lambda) = \mathcal{N}(\xi; \mu, \Sigma)$ (ie. a MVN distribution with the parameters $\lambda=(\mu, \Sigma)$). One defines $S_\lambda$ such that $S_\lambda(\xi)=\zeta$, eg. $\zeta = L^{-1}(\xi-\mu)$ where $L$ is the Cholesky decomposition of $\Sigma=LL^T$. The Jacobian of this distribution is 1 (volume conservation), so the cost function reads
\begin{multline}
-\mathcal{L}(\lambda) = \underbrace{\mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left[ \log p(T^{-1}(S_\lambda^{-1}(\zeta)),\mathcal{D}) + \log |J_{T^{-1}}(S_\lambda^{-1}(\zeta))| \right]}_{model} \\ + \underbrace{\mathbb{H}[q(\xi;\lambda)]}_{guide}
\label{eq-loss-svi-3}
\end{multline}
where $\mathbb{H}(q)$ is the Shannon entropy of the MVN distribution which is independent of the model, so its gradient can be computed once for all and reused in any user model.  
For the use case of the MVN (auto)guide, one gets
\begin{align}
-\nabla_\mu \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)]
+ \nabla_\xi \log|J_{T^{-1}}(\xi)|
\right\} \nn \\
-\nabla_L \mathcal{L} &= \mathbb{E}_{\zeta\sim \mathcal{N}(0,I)}\left\{
\left[ \nabla_z \log p(z,\mathcal{D}) \times \nabla_\xi[T^{-1}(\xi)] \right.\right. \nn\\
&\left. \left. \qquad\qquad\qquad + \nabla_\xi \log|J_{T^{-1}}(\xi)| \right] \times \xi^T 
\right\}+ (L^{-1})^T
\end{align}
\FrL{Yeah. this is way TMI, and also goes against the idea that with autodiff we don't need to manually work out all these details.} \FrL{let's move the MVN details to an appendix.}
The expectations used in the above expressions are computed with $\zeta$ i.i.d samples from $\mathcal{N}(0,1)$ distribution which reveals the ``S'' of SVI. As one can imagine all the jacobian computations take benefit of the automatic differentiation offers by JAX.

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%\section*{Bibliography}
%\bibliographystyle{elsarticle-harv} 
\bibliographystyle{aa} 
\typeout{}
\bibliography{refs}
